At this stage of the strategy, the only place that the program counter
is used is when the first method is started, when it is used to select
the method action to execute, which will then proceed without any need
for the program counter value.
This can be eliminated by replacing it with a choice over the method
rather than the program counter.
This must be performed in the two places that the $Running$ action
occurs:~the $MainThread$ and $NotStarted$ actions.
Since the main action of $Thr$ is a guarded choice of these actions
depending on whether its thread parameter is the $main$ thread, these
actions may be thought of as two alternative main actions for $Thr$.

The context of the $Running$ action in both of these main actions is
the same:~the frame stack has only one frame and the program counter
is set to the entry point of a method.
This means that the same rule can be used for both main actions by
introducing an assumption that states that context.
Additionally, we can use the fact that each method action will, when
started with a frame stack containing a single frame, cause the frame
stack to become empty.
This allows us to eliminate the loop in $Running$, reducing it
entirely to a choice of method action from the class and method
identifier.
The overall transformation of $Running$ in its context is described by
Rule [\nameref{main-action-refinement-rule}].
\begin{restatable}[Main Action Refinement]{crule}{MainActionRefinementRule}
  \label{main-action-refinement-rule}
  If $entry(c_i,m_i) = j_i$ for $i \in \{1, \dots, n\}$ and
  \begin{circus}
    \{ \# frameStack = 1 \} \circseq M_i \\
    {} = {} \\
    \{ \# frameStack = 1 \} \circseq M_i \circseq \{ frameStack = \emptyset \}
  \end{circus}
  \setlength{\zedindent}{0.25cm}
  \begin{circus}
    \begin{array}{l}
      \{ \# frameStack = 1 \\
      \t1 {} \land pc = entry(cid, mid) \} \circseq \\
      \circmu X \circspot \\
      \t1 \circif frameStack = \emptyset \circthen \Skip \\
      \t1 {} \circelse framestack \neq \emptyset \circthen {}  \\
      \t2 \circif pc = j_1 \circthen M_1 \\
      \t2 \cdots \\
      \t2 {} \circelse pc = j_n \circthen M_n \\
      \t2 \circfi \circseq Poll \circseq X \\
      \t1 \circfi
    \end{array}
    \circrefines_A
    \begin{array}{l}
      \{ \# frameStack = 1 \} \circseq \\
      \circif (cid, mid) = (c_1, m_1) \circthen M_1 \\
      \cdots \\
      {} \circelse (cid, mid) = (c_n, m_n) \circthen M_n \\
      \circfi \\
    \end{array}
  \end{circus}
\end{restatable}
Once this has been performed, the program counter value is no longer
used to determine the control flow of the program, so a trivial data
refinement can be performed to eliminate $pc$ from the state of the
$Thr$ process.