\documentclass[a4paper,10pt]{report}

\usepackage[in]{fullpage}
\usepackage{parskip}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{pgfgantt}
\usepackage{amsmath}
\usepackage[color,cntglobally]{circus}
\usepackage[backend=bibtex,style=numeric-comp,sorting=nyt,sortcites=true,maxnames=4]{biblatex}
\usepackage[section]{placeins}
\usepackage{fixltx2e}
\usepackage{rotating}
\usepackage{pdfpages}

%\newcommand{\Circus}{{\sf\slshape Circus}}

\title{A Framework for Verifying Safety-Critical Java Virtual Machines
with Ahead-of-time Compilation \\ \hfill \\
\Large Thesis Outline}
\author{James Baxter}
\date{}

\bibliography{../Thesis/literature.bib} 
\bibliography{draftthesis.bib}

\let\olditem\item
\renewcommand{\item}[1][]{\olditem{\bfseries #1}}

%TC:group zed 0 displaymath
%TC:group axdef 0 displaymath
%TC:group schema 1 displaymath
%TC:group circus 0 displaymath
%TC:group circusaction 0 displaymath
%TC:macroword \Circus 1 

\begin{document}
\maketitle

\begin{abstract}
  In recent years Java has been increasingly considered as a language
  for safety-critical embedded systems.
  However, some features of Java are unsuitable for such systems and
  this has resulted in the creation of Safety-Critical Java (SCJ).
  The different scheduling and memory management model of SCJ means
  that a specialised virtual machine is required to run SCJ programs.
  Given the safety-critical nature of the applications, it must be
  ensured that the virtual machine is correct, but so far no SCJ
  virtual machine has been formally verified.
  We propose a framework for verification of SCJ virtual machines
  comprising a formal model of requirements and a compilation strategy
  from Java bytecode to native code together with mechanised proofs of
  correctness.
  In this report, we describe the structure of the thesis that will be
  written for the work and provide a plan for the remainder of the
  work.
\end{abstract}

\tableofcontents

\chapter{Introduction}

% short explanation of chapter: motivation, objectives, document
% structure
In this chapter, we present a present a description of the motivation
for and objectives of our work.
The overall scope of our research has not changed substantially since
the progress report so this chapter simply presents a brief
introduction to our work.
This chapter begins with a description of the motivation of our work
in Section~\ref{motivation-section}, followed by an explanation of the
overall objectives of our work in Section~\ref{objectives-section}.
The structure of the remainder of this report is then described in
Section~\ref{document-structure-section}.


\section{Motivation}
\label{motivation-section}

The Java programming language~\cite{gosling2013} has achieved great
popularity and is now used in a wide variety of areas.
One particular area of interest in which Java is used is that of
embedded systems, where it was realised that the features of
portability, modularity, safety and security that Java offers could be
of use~\cite{mulchandani1998}.
As Java programs are compiled to be run by the Java Virtual Machine
(JVM), this required creating JVMs for embedded devices and, indeed,
research has gone into making smaller and smaller JVMs to accommodate
a wider range of embedded devices~\cite{caska2011,thomm2010}.

However, many embedded systems have real-time requirements, meaning
that strict timing properties are required to be adhered to.
Features of Java such as the garbage collector and its concurrency
model make it difficult to ensure such properties, thus making
standard Java unsuitable for real-time systems.
To address this problem, the Real-Time Specification for Java
(RTSJ)~\cite{gosling2000} was created.
The RTSJ extends Java with some additional features that allow for
greater time predictability.
One feature the RTSJ adds is scoped memory, allowing memory to be
allocated for use in a particular scope and deallocated when that
scope is left, thus allowing for prediction of when memory
deallocations occur.
Another feature of the RTSJ is a more predictable priority scheduler
that better allows the execution time of higher priority threads to be
computed.

Though the RTSJ provides for real-time considerations, many real-time
embedded systems require certification under safety standards such as
\mbox{DO-178C} and ISO~26262.
To facilitate the certification of such systems, Safety-Critical Java
(SCJ)~\cite{locke2013} was created.
SCJ is based on the RTSJ but removes the features of the RTSJ that
would make it difficult to certify programs written using it.
Some particular differences of SCJ from standard Java are in the areas
of scheduling and memory management.

SCJ's scheduling model is based around organising a program into a
series of missions, which are executed sequentially in an order chosen
by a programmer-supplied mission sequencer.
Each mission may contain multiple asynchronous event handler threads,
which are created when the mission is initialised and run throughout
the duration of the mission.
The event handlers within a mission are scheduled according to a
priority scheduling system, in which the running handler is always one
of the released handlers with the highest priority so that high
priority handlers are not sharing processor time with low priority
handlers.
Once the mission is requested to terminate, the event handler threads
are stopped, the mission's resources are cleaned up, and the mission
sequencer selects the next mission.

The memory management of SCJ makes use of the scoped memory areas of
RTSJ, adapted for the SCJ mission model.
Memory is allocated in memory areas, which have varying lifetimes.
Each mission has a memory area associated with it that persists for
the duration of the mission, while each event handler within the
mission has a memory area that persists only during releases of the
handler.
Additional nested private memory areas can be created, which can be
entered and left as needed, being cleared when they are left.
There is also an immortal memory area, in which memory is never freed
and persists for the entire duration of the program.
A system of annotations allows for static checking that dangling
pointers cannot arise from misuse of the scoped memory
model~\cite{tang2010}.

Due to the safety-critical nature of the software SCJ is designed to
create, it is necessary to show that it is correct.
While the design of SCJ allows for it to be more easily shown that
programs will execute correctly, greater certainty and precision can
be obtained through the use of formal methods.
There has already been work on generating correct SCJ programs from
formal specifications~\cite{cavalcanti2011, cavalcanti2013}, which
ensures that an SCJ program will have the behaviour intended.
Some work has also been done on formalising the SCJ memory
model~\cite{cavalcanti2011a}, allowing its correctness to be ensured.

However, although SCJ can be compiled by a standard Java compiler, SCJ
requires a specialised virtual machine due to the differences between
SCJ and standard Java.
There have been several SCJ virtual machines (SCJVMs) created,
including the icecap HVM~\cite{sondergaard2012}, Fiji
VM~\cite{pizlo2009}, OVM~\cite{armbruster2007},
HVM\textsubscript{TP}~\cite{luckow2014} and PERC Pico~\cite{atego2015,
  richard2010}.
Of these, only Fiji VM and icecap HVM appear to be maintained and,
while icecap provides support for most of SCJ, Fiji VM does not
necessarily provide proper support for all aspects of SCJ.
Since the correct execution of an SCJ program depends on the
correctness of the underlying virtual machine, it must be ensured in
addition to checking the correctness of the program.
While there has been work on using formal methods to show SCJ program
correctness, no SCJVM has been verified.

Additionally, having to execute a program via a virtual machine
presents a cost in terms of memory and time resources, which may not
be available on the embedded real-time systems SCJ targets.
For this reason, all known SCJVMs, including the ones listed above
compile the Java bytecode to native code ahead-of-time.
This can be particularly seen in the examples of Fiji VM and the
icecap HVM, which compile Java bytecode to C code.
The correctness of this compilation to native code must also be
ensured, in addition to the correctness of the infrastructure
supporting the running program.
So the problem of ensuring the correctness of an SCJVM is, in part, a
compiler verification problem.

There has already been much research in the area of compiler
correctness, with most of the research following one of two
approaches.
The majority of the literature on compiler correctness follows a
commuting-diagram approach, in which the compilation function is shown
to commute with functions defining the semantics of the source and
target languages.
This approach was first identified by Lockwood
Morris~\cite{morris1973} and later refined by Thatcher \emph{et
  al.}~\cite{thatcher1979}, but can be seen in much of the earlier
work, including the earliest work by McCarthy and
Painter~\cite{mccarthy1967}.
The commuting-diagram approach has also been used in more recent work,
including some very comprehensive work as part of the CompCert
project~\cite{leroy2009a, leroy2009b}.
As this approach is based on the use of ordinary functions to describe
compiler correctness, it can be readily encoded in an automated
theorem prover and there have been several works that have made use of
automated theorem provers~\cite{klein2006, milner1972, nipkow2000}.

A recent variation of the commuting-diagram approach is based on an
operational semantics of the source language~\cite{bahr2015}.
In this work, the operational semantics of the source language and a
way of relating the source and target semantics are used to derive a
different operational semantics of the source language acting on the
state of the target machine.
The semantics of the target language are then identified as part of
that operational semantics and it is transformed to extract a
compilation function.
This approach may be viewed as variant of the commuting-diagram
approach in which the compilation function is derived from the source
and target semantics and the relationship between them, rather than
being verified by those elements of the commuting-diagram.

The second main approach to compiler verification is the algebraic
approach proposed by Hoare~\cite{hoare1991} and implemented by
Sampaio~\cite{hoare1993, sampaio1993}.
The algebraic approach is based around the notion of refinement, which
formally captures the idea of a program being an implementation of a
less deterministic specification.
In the algebraic approach, the source and target languages are defined
in the same semantic space and proven refinement laws are used to
refine the source program to a normal form representing the target
machine running the target code.
This approach has the advantage that the rules used to perform the
compilation are known to be correct, thus making the resultant
compiler correct by construction.
It also allows for passing seamlessly from the source language to the
target language using algebraic laws, since the source and target
languages are in the same semantic space.
The algebraic approach has not been used particularly widely but, in
addition to Sampaio's work, there has been work using the algebraic
approach for compilation of object-oriented languages~\cite{duran2005,
  duran2010} and for hardware compilation~\cite{perna2010, perna2011}.

\section{Objectives}
\label{objectives-section}

Since there appears to be no formally verified SCJVM and formal
verification of an SCJVM is desirable, we propose to create a
framework for formal verification of an SCJVM.
This framework will consist of the following parts:
\begin{itemize}
\item A specification of the services required to be provided by an
  SCJVM,
\item A compilation strategy from Java bytecode to C,
\item A formal model of the specification and compilation strategy,
\item Proofs of the correctness of the formal model, and
\item A mechanisation of the proofs in an automated theorem prover.
\end{itemize}
Each of these parts will be discussed in what follows.

First, it is necessary to have an explicit account of what is required
of an SCJVM.
Even though the ultimate aim is one of formal verification, an
informal specification is still needed to guide the development of the
formal model.
There is currently no clear specification for SCJVMs.

It should be noted that the role of an SCJVM is not merely to execute
bytecode instructions.
An SCJVM must also provide services to support the SCJ API in areas
such as scheduling and memory management.
An informal specification of all aspects of the SCJVM is required to
support creation of a formal model.
On the other hand, the semantics of SCJ bytecode does not differ much
from that of standard Java bytecode, so much of the Java Virtual
Machine specification~\cite{lindholm2014} also applies to SCJVMs.

The need for a compilation strategy arises from the fact that most
existing SCJVMs compile Java bytecode to some native code in order to
improve performance on embedded systems.
We will focus on compilation from Java bytecode to C for much the same
reasons as the icecap HVM and Fiji VM:~C is a language already widely
used in embedded systems and is sufficiently low level to enable its
use as an efficient target language while retaining enough abstraction
to be used by the programmer.

The specification and compilation strategy must be formalised.
As noted previously, there are two main approaches to formalising
compilers and showing their correctness: the commuting-diagram
approach and the algebraic approach.
We will follow the algebraic approach in our work as it relates the
source and target languages in a clearer way than the
commuting-diagram approach.
Additionally, the commuting-diagram approach requires the definition
of additional functions to relate the source and target semantics,
thus relying more on definitions, which may be incorrect, than on
mathematically derived results.
The approach of deriving a compiler from an operational semantics
could be of use as it does not have this issue.
That approach, however, relies on knowledge of the state of the target
machine and a way of identifying the semantics of the target code
within the compilation function.
This is not easily done with a complex language like C, though it
works well for machine languages.

We require a formal language in which to specify the source and target
languages, as well as for specifying the virtual machine services.
For this we will use the \Circus{} specification
language~\cite{oliveira2009}, which is based on Z
notation~\cite{woodcock1996} and CSP~\cite{roscoe2011}.
\Circus{} uses CSP to specify processes that communicate over channels
and uses Z notation to specify state and data operations that can be
encapsulated in processes.
The reason for the choice of \Circus{} as our formal notation is due
to the fact that \Circus{} is designed as a notation for refinement
and so works well with the algebraic approach.
The combination of Z notation and CSP also gives far greater
expressive power than using those languages separately.
Finally, the use of \Circus{} helps to tie this work into the existing
specification work surrounding SCJ~\cite{cavalcanti2011,
  cavalcanti2011a, cavalcanti2013, zeyda2011}.

The consistency of the specification and the correctness of the
compilation strategy must be proved by formal mathematical proof from
algebraic laws known to be correct.
\Circus{} has many such laws already developed and has an underlying
semantics given using the model of Hoare and He's Unifying Theories of
Programming (UTP)~\cite{hoare1998} that can be used to verify the
correctness of the laws.

To ensure the laws are understood and applied correctly, it is helpful
to mechanise the proofs in an automated theorem prover.
There are a variety of tools for machine checking of \Circus{}
specifications.
The \Circus{} parser and typechecker included as part of the Community
Z Tools~\cite{malik2011, xavier2008, malik2005, miller2005} can be
used to perform basic checking of a \Circus{} specification and
integration with the Z/Eves theorem prover~\cite{saaltink1997} can be
used to prove the Z portions of the \Circus{} specification.
Process properties such as deadlock-freedom can be checked by
translating the \Circus{} specification to CSP and checking with a
tool such as FDR~\cite{gibson-robinson2014}.
For more comprehensive proving of properties about \Circus{}
specifications, recent work on Isabelle/UTP~\cite{foster2015} can be
used to permit reasoning about \Circus{} in the Isabelle theorem
prover~\cite{nipkow2002}.

Altogether, our objective is a framework for verification of SCJVMs
consisting of mechanised formal models together with proofs of
properties about them.
These formal models will cover both the services that must be provided
by a running SCJVM and a compilation strategy for translating Java
bytecode to native code.
With our results, SCJVM developers will be able to create provably
correct ahead-of-time compiling SCJVM implementations and check the
correctness of those implementations.


\section{Document Structure}
\label{document-structure-section}

The remainder of this report proceeds with an explanation of the
structure of the thesis in Chapter~\ref{thesis-structure-chapter},
describing each thesis chapter in a separate section.
This is then followed by a description of the plan for the remainder
of the research in Chapter~\ref{plan-chapter}.
 

\chapter{Thesis Structure}
\label{thesis-structure-chapter}

In this chapter, we describe the proposed structure of the thesis that
will be written.
Each section in this chapter corresponds to a chapter in the thesis
and in it we detail the contents of the corresponding thesis chapter.
We also discuss the progress of the research documented in each
chapter of the thesis.

\section{Introduction}

The first chapter of the thesis will be an introductory chapter, in
which we will lay out the main problem solved in the thesis.
We describe its relevance, the overall objectives pursued in the work,
and the structure of thesis.
As this will be an introductory chapter, no specific research is
required for it.
The content of this chapter has already been written, primarily as
part of writing previous deliverables over the course of the work.
It should only require some minor revision for the final thesis.
The material in the previous chapter of this document is very close to
the current draft of the introduction of the thesis.

\section{Compilers and Virtual Machines for Java-like languages in the
  Safety-critical Domain}

In the second thesis chapter we will review the literature relevant to
our work and explain various preliminary concepts that must be
introduced before the work in the thesis can be fully understood.
The only research that needs to be performed for this chapter is the
literature review, which has already been completed prior to beginning
the bulk of the work.
As with the introduction, the content for this chapter has already
been written up and will only require minor changes to take into
account new developments in the literature.
We refer to~\cite{baxter2016draftthesis}~for the current contents of
this chapter.
It is structured with the following sections:
\begin{enumerate}
\item[Java for Safety-critical Systems], in which we describe the
  history of SCJ and explains why a specialised variant of Java is
  necessary for safety-critical systems.
\item[Safety-Critical Java], in which we describe SCJ itself, including the
  fact that SCJ is divided into three compliance levels, the mission
  model of SCJ, and the memory model of SCJ.
\item[Virtual Machines for Safety-Critical Java], in which we consider
  some of the existing SCJVM implementations, looking at their
  approaches to running SCJ programs, the extent to which they
  implement the SCJ specification, and whether they are still
  maintained, and we note that no formal verification work has
  been performed on any existing SCJVM.
\item[Compiler Correctness], in which we discuss different methods of
  verifying compilation, in particular the commuting-diagram approach
  and the algebraic approach, which are the two main approaches.
\item[\Circus{}], in which we describe the \Circus{} specification
  language so that the presentation of the model in subsequent
  chapters can be understood.
\item[Final Considerations], in which we summarise and consider the
  contents of the chapter and conclude that we should adopt the
  algebraic approach to verifying the compilation, with \Circus{} as
  the specification language.
\end{enumerate}

\section{Safety-Critical Java Virtual Machine Services}

This will be the first technical chapter of the thesis and will cover
the first part of the SCJVM model.
It consists of the services that must be offered by an SCJVM to
support the running of an SCJ program but will not be directly
affected by the compilation strategy.
The work that needs to be carried out for this chapter is the writing
of the SCJVM services model and that has largely already been
completed, though further changes to the model may be required to
comply with changes to the developing SCJ specification.
This chapter has already been written up, with the informal
explanation of the SCJVM services based on material from a paper we
published at the 13th International Workshop on Java Technologies for
Real-time and Embedded Systems~\cite{baxter2015a}, which described an
earlier version of the model.
The model has been explained in detail
in~\cite{baxter2016draftthesis}, but the existing material for the
chapter will need to be revised to make it consistent with more recent
changes to the model and present it in a briefer style.
Its envisaged structure of sections is as follows:
\begin{enumerate}
\item[Overview], in which we describe the overall structure of the
  SCJVM services and how it relates to the rest of the SCJVM.
\item[Memory Manager API], in which we give an informal description of
  the services related to memory management that are specified in the
  model.
\item[Scheduler API], in which we give an informal description of the
  services related to scheduling that are specified in the model.
\item[Real-time Clock API], in which we give an informal description
  of the services related to the real-time clock and timekeeping that
  are specified in the model.
\item[Formal Model], in which we explain the formal model of the SCJVM
  services, with each part of the model explained in a separate
  subsection.
\item[Final Considerations], in which we summarise the contents of the
  chapter, explain the relationship between the different parts of the
  model, and discuss the progress in subjecting the model to proof.
\end{enumerate} 

\section{Core Execution Environment}

Following the explanation of the SCJVM Services model in the previous
thesis chapter, the Core Execution Environment chapter will detail the
part of the SCJVM model that will be affected by the compilation
strategy, including execution of the code and handling of objects.
The code execution includes both interpretation of the bytecode and
execution of the C program that results from compilation of the
bytecode.
The model of the bytecode interpreter and the structure of the C code
that the compilation strategy will generate have already been created.
However, due to the interaction of this part of the model with the
compilation strategy, it may require some changes as development of
the compilation strategy continues.
The bytecode interpreter model has been written up but must be changed
to reflect changes in the model and include the C code model.
The structure of this chapter is as follows:
\begin{enumerate}
\item[Overview], in which we describe the overall purpose of the core
  execution environment, its structure, and the different approaches
  of interpretation and compilation to C.
\item[Bytecode Interpreter Model], in which we describe each of the
  components of our model for the core execution environment with an
  interpreter.
\item[C Code Model], in which we describe the structure and
  representation of the C code that will be generated from compilation
  of the bytecode.
\item[Final Considerations], in which we summarise the contents of the
  chapter, and discuss some interesting points about the model and its
  possible uses.
\end{enumerate}

\section{Compilation Strategy}

This thesis chapter will describe the strategy for transforming the
core execution environment model loaded with a given SCJ program to an
equivalent C program of the form described in the previous thesis
chapter.
Then we will provide an overview of the compilation strategy,
describing each of the stages and stating the main compilation theorem
that the strategy proves, which is that the model can be transformed
to a particular form representing a valid C program.
The detailed rules that must be applied at each stage of the
compilation strategy will then be described and we will explain how
the bytecode program is transformed into a C program through provably
correct transformations.

We are part of the way through the research for this chapter, with a
rough outline of the compilation strategy already drawn up.
The next stage of the research will be to work out detailed
compilation rules for each stage of the compilation strategy and show
that those rules are mathematically correct.
None of the material for this chapter has been written up yet.
The planned structure for the chapter is as follows:
\begin{enumerate}
\item[Overview], in which we describe the overall compilation strategy
  and list the major steps of the compilation, which are defined in
  detail in later sections of the chapter.
\item[Elimination of Program Counter], in which we explain the
  compilation rules necessary to remove dependence on an explicit
  program counter and introduce control structures.
\item[Elimination of Stack], in which we explain how information from
  the Java stack is converted to variables and parameters in C in
  order to remove dependence on the stack.
\item[Collapsing of Interpreter and Launcher], in which we describe
  how the interpreter and launcher components of the model are merged
  in order to remove the interpreter's state.
\item[Data Refinement of Memory], in which we describe how the
  representation of objects is introduced via a data refinement.
\item[Final Considerations], in which we summarise the compilation
  strategy and discuss any additional considerations that could not be
  addressed when explaining the strategy itself.
\end{enumerate}

\section{Evaluation}

The focus of this chapter will be evaluation of the compilation
strategy.
This will be performed by giving some examples of how the
compilation strategy acts on the bytecode produced from real SCJ
programs, so that it can be seen that equivalent C code is generated
by the compilation strategy.
The C code that results from application of the compilation strategy
will be compared to code generated by the icecap HVM as that is the
most developed publicly available implementation of SCJ.

Mechanisation of the correctness proofs for the compilation strategy
will also be considered in this chapter, since that provides greater
certainty that the strategy is valid.
We will describe the basic principles of the mechanisation and the
discuss the results obtained from it in terms of evaluating the
correctness of the compilation strategy.

The work for this thesis chapter remains to be done.
Some of the work of developing examples of the application of the
compilation strategy will be done during the creation of the
compilation strategy, but it is unlikely that work will done on the
mechanisation until after the full compilation strategy has been
constructed.
As the research for this chapter has not yet been completed, nothing
has been written for it yet.
The structure of this chapter is expected to be as follows:
\begin{enumerate}
\item[Examples], in which we illustrate the use of the compilation
  strategy by applying it to several examples of small SCJ programs.
\item[Mechanisation], in which we show how the compilation strategy
  and the proofs of its correctness can be mechanised.
\item[Final Considerations], in which we summarise and discuss the
  overall evaluation of the compilation strategy.
\end{enumerate}

\section{Conclusions}

The final chapter of the thesis will be the Conclusions chapter, in
which we will summarise the contributions of the thesis, including
discussing how it builds upon the existing work in the field.
We will also discuss some possible directions for future work outside
the scope of the thesis.
Most of the material for this chapter has already been written, but
more will be written as progress is made on the rest of the thesis.
Its envisaged structure is standard:
\begin{enumerate}
\item[Summary of Contributions], in which we summarise the
  contributions of the thesis in the context of existing work.
\item[Future Work], in which we discuss possible future directions for
  the work beyond the thesis.
\end{enumerate}

\chapter{Plan}
\label{plan-chapter}

In this chapter we discuss the plan for the remainder of the research.
The tasks still to be completed for each thesis chapter have been mentioned
in the previous chapter of this document.
The proposed schedule for completing these tasks is much the same as
in the progress report delivered earlier this year and we believe the
work is currently on schedule.
The tasks left to be completed are as follows, in the order they will
be completed:
\begin{itemize}
\item[Develop compilation strategy]: We will continue to develop the
  rough outline of the compilation strategy that we have at present
  into a more detailed compilation strategy.
\item[Hand proofs of compilation rules]: When we have fully developed
  the compilation strategy, we will show the rules tha make up the
  compilation strategy can be constructed from know laws of \Circus{},
  thus ensuring its correctness.
\item[Mechanise proofs of compilation rules]: We will then mechanise
  the proofs of the correctness of the compilation rules in order to
  gain greater certainty of their correctness.
\item[Implementation and example development]: After completing and
  proving the compilation strategy, we will develop further examples
  and consider automated application of the compilation rules in order
  to evaluate the strategy.
\item[Writing up results]: Finally, while some writing up of results
  will occur during the work, some time has been set aside at the end
  to complete the the writing up of results.
\end{itemize}
The expected schedule for these tasks is shown in
Figure~\ref{ganttchart}.

\begin{sidewaysfigure}
\centering
\begin{minipage}{\textheight}
\centering
\begin{ganttchart}[
  vgrid,
  x unit=0.7cm,
  time slot format = isodate-yearmonth,
  compress calendar
  ]{2016-07}{2017-12}
  \gantttitlecalendar{year, month=shortname} \\
  \ganttbar{Develop compilation strategy}{2016-07}{2016-08} \\
  \ganttbar{Hand proofs of compilation rules}{2016-08}{2017-01} \\
  \ganttbar{Mechanise proofs of compilation rules}{2017-01}{2017-03} \\
  \ganttbar{Implementation and example development}{2017-04}{2017-05} \\
  \ganttbar{Writing up results}{2016-08}{2016-08}
  \ganttbar{}{2017-01}{2017-01}
  \ganttbar{}{2017-06}{2017-09} \\
\end{ganttchart}
\end{minipage}
\caption{A Gantt chart showing the schedule for the remainder of the
  planned research}
\label{ganttchart}
\end{sidewaysfigure}

\printbibliography

\end{document}