\chapter{Compilation Strategy}
\label{strategy-chapter}

In this chapter we describe our compilation strategy for refining SCJ
bytecode to C code.
We begin in Section~\ref{compilation-overview-section} with an
overview of our compilation strategy.
Then, in Section~\ref{compilation-assumptions-section} we describe the
requirements on the source program for the compilation strategy to be
applied.
Afterwards, we describe each stage of the strategy in a separate
section.
The first stage, which we call \emph{Elimination of Program Counter},
is described in Section~\ref{elimination-of-program-counter-section}.
The second stage, called \emph{Elimination of Frame Stack}, is
described in Section~\ref{elimination-of-frame-stack-section}.
Finally, the third stage of the strategy, which is called \emph{Data
  Refinement of Objects}, is described in
Section~\ref{data-refinement-of-objects-section}.
We then show how the stages fit together to show the compilation as a
whole to be correct, in Section~\ref{main-theorem-proof-section}, and
conclude with some final considerations in
Section~\ref{compilation-final-considerations-section}.

\section{Overview}
\label{compilation-overview-section}

Our compilation strategy refines the $CEE(bc,cs,instCS,sid,initOrder)$
process defined in Section~\ref{cee-interpreter-section} to obtain the
$CCEE_{bc,cs}(sid, initOrder)$ process in
Section~\ref{cee-c-code-section}.
The overall theorem for the strategy, and, therefore, the main result
presented in this chapter, is as follows.
\begin{thm}[Compilation Strategy]\label{main-theorem}
  Given $bc$, $cs$ and $sid$, there are processes $StructMan_{cs}$ and
  $CProg_{bc,cs}$ such that,
  \setlength{\zedindent}{0.5cm}
  \begin{circus}
    CEE(bc,cs,instCS,sid,initOrder) \circrefines StructMan_{cs} \parallel
    CProg_{bc,cs} \parallel Launcher(sid, initOrder).
  \end{circus}
\end{thm}
$StructMan_{cs}$ manages objects represented by C structs that
incorporate the class information from $cs$, refining the process
$ObjMan$, which handles abstract objects.
$CProg_{bc,cs}$ refines the $Interpreter$, with the $Thr$ processes
refined into the $CThr_{bc,cs}$ processes described in
Section~\ref{cee-c-program-subsection}.
This means that the threads from SCJ are mapped onto threads in C,
since we do not dictate a particular thread switch mechanism in either
the source or target models.

The compilation strategy is split into three stages.
Each stage has a theorem describing it, for which the strategy acts as
a proof.
The proof of Theorem~\ref{main-theorem}, presented in
Section~\ref{main-theorem-proof-section}, is obtained by an
application of the theorems for each stage.
Each stage of the compilation strategy handles a different part of the
$Interpreter$ state:~the $pc$, the $frameStack$, and objects.
They operate over each of the $Thr$ processes, managed by the SCJVM
services.

The first stage, \emph{Elimination of Program Counter}, introduces the
control constructs of the C code.
This removes the use of $pc$ to determine the control flow of the
program.
The choice over $pc$ values is replaced with a choice over method
identifiers pointing to sequences of operations representing method
bodies.

In the second stage, \emph{Elimination of Frame Stack}, the
information contained on the $frameStack$, which is the local variable
array and operand stack for each method, is introduced in the C code.
This is done by introducing variables and parameters to represent each
method's local variables and operand stack slots.
A data refinement is then used to transform each operation over the
$frameStack$ to operate on the new variables.
The $frameStack$ is then eliminated from the state.

In the final stage, \emph{Data Refinement of Objects}, the class
information from $cs$ is used to create a representation of C structs.
This means that $ObjMan$, which has a very abstract representation of
objects, is transformed into $StructMan$.
The operations on objects are then changed to access the structs for
the objects in a more concrete way that represents the way struct
fields are accessed in C code.

% This yields final method actions of a form similar to that of the
% example shown below, which is taken from the \texttt{InputHandler}
% presented in Section~\ref{model-section}.
% \begin{circusaction}
%   InputHandler\_HandleAsyncEvent \circdef \\
%   \t1 \circval var0 \circspot \circvar var1, stack0, stack1 : Word \circspot \\
%   \t1 stack0 := var0 \circseq Poll \circseq getObject!stack0 \then getObjectRet?struct \\
%   \t1 {} \then stack0 := (castInputHandler~struct).input \circseq \dots
% \end{circusaction}
% The \texttt{handleAsyncEvent()} method of \texttt{InputHandler} is
% compiled to the action $InputHandler\_HandleAsyncEvent$, with the
% implicit \texttt{this} parameter represented as a value parameter
% $var0$.
% The local variable ($var1$) and stack slots ($stack0$ and $stack1$)
% are represented as \Circus{} variables.
% The operations of the C code are composed in sequence, with an action
% named $Poll$ that polls for thread switches present at the points
% where thread switches may occur. 
% Stack operations are represented as assignments. 
% For instance, $stack0 := var0$ arises from the compilation to load a
% local variable into a stack slot.
% Access to objects is performed by communicating with $StructMan_{cs}$
% to obtain the struct for the object, then casting it to the correct
% type, and accessing the required value.
% Above, we obtain the value of the $input$ field from an $InputHandler$
% object.
% The communication with $StructMan_{cs}$ is performed via the
% $getObject$ channel and the function $castInputHandler$ is used to map
% the $ObjectStruct$ returned from the communication to a type
% representing an object of \texttt{InputHandler}.

\section{Assumptions about source bytecode}
\label{compilation-assumptions-section}

\input{compilationstrategy/compilationassumptions.tex}

\section{Elimination of Program Counter}
\label{elimination-of-program-counter-section}

\input{compilationstrategy/epc-overview.tex}

\subsection{Running Example}
\label{overview-subsection}

\input{compilationstrategy/epc-runningexample.tex}

%\FloatBarrier

\subsection{Expand Bytecode}
\label{expand-bytecode-subsection}

\input{compilationstrategy/epc-expandbytecode.tex}

\subsection{Introduce Sequential Composition}
\label{introduce-forward-sequence-subsection}

\input{compilationstrategy/epc-sequentialcomposition.tex}

\subsection{Introduce Loops and Conditionals}
\label{introduce-loops-and-conditionals-subsection}

\input{compilationstrategy/epc-loopsandconditionals.tex}

\subsection{Resolve Method Calls}
\label{resolve-method-calls-subsection}

\input{compilationstrategy/epc-methodcalls.tex}

\subsection{Refine Main Actions}
\label{refine-main-actions-subsection}

\input{compilationstrategy/epc-mainactions.tex}

\subsection{Remove \texorpdfstring{$pc$}{pc} From State}
\label{remove-pc-from-state-subsection}

\input{compilationstrategy/epc-pcelimination.tex}

\section{Elimination of Frame Stack}
\label{elimination-of-frame-stack-section}

\input{compilationstrategy/efs-overview.tex}

\subsection{Remove Launcher Returns}
\label{remove-launcher-returns-subsection}

\input{compilationstrategy/efs-removelauncherreturns.tex}

\subsection{Localise Stack Frames}
\label{localise-stack-frames-subsection}

\input{compilationstrategy/efs-localisestackframes.tex}

\subsection{Introduce Variables}
\label{introduce-variables-subsection}

\input{compilationstrategy/efs-introducevariables.tex}

\subsection{Remove \texorpdfstring{$frameStack$}{frameStack} From
  State}
\label{remove-frameStack-from-state-subsection}

\input{compilationstrategy/efs-framestackelimination.tex}

\section{Data Refinement of Objects}
\label{data-refinement-of-objects-section}

\input{compilationstrategy/dro-overview.tex}

\section{Proof of Main Theorem}
\label{main-theorem-proof-section}

\input{compilationstrategy/main-theorem-proof.tex}

\section{Final Considerations}
\label{compilation-final-considerations-section}

\input{compilationstrategy/final-considerations.tex}