\documentclass[a4paper,11pt,twoside,openright]{report}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[hidelinks]{hyperref}
\usepackage[color,cntglobally]{circus}
\usepackage[table]{xcolor}
\usepackage{array}
\usepackage[in]{fullpage}
\usepackage{tikz}
\usepackage{parskip}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[backend=bibtex,style=numeric-comp,sorting=nyt,sortcites=true,maxnames=4,defernumbers=true]{biblatex}
\usepackage{placeins}
\usepackage{mathdots}
\usepackage{multicol}
\usepackage[savemem]{listings}
\usepackage{vwcol}
\usepackage{fixltx2e}
\usepackage{comment}
\usepackage{subcaption}
%\usepackage{cleveref}
\usepackage{thmtools}
\usepackage{thm-restate}
\usepackage{supertabular}
\usepackage{afterpage}
\usepackage{letltxmacro}
\usepackage[normalem]{ulem}
\usepackage{tikz-uml}
\usepackage{titling}
\usepackage{lscape}
\usepackage{pdflscape}
\usepackage{needspace}
\usepackage[explicit]{titlesec}

\makeatletter
\renewcommand{\boolean}{\zordop{\csym@boolean}}
\makeatother

\CountDefsfalse

\definecolor{black}{rgb}{0,0,0}
\definecolor{ZedBoxColor}{rgb}{0,0,0}
\definecolor{AnnotationColor}{rgb}{0,0,0}
\definecolor{ZedColor}{rgb}{0,0,0}

\definecolor{headcol}{rgb}{0,0,0} 

\newcommand{\resetZColors}{%
  \colorlet{ZedBoxColor}{black}%
  \colorlet{AnnotationColor}{black}%
  \colorlet{ZedColor}{black}%
}

\newcommand{\setZColors}[1]{%
  \colorlet{ZedBoxColor}{#1}%
  \colorlet{AnnotationColor}{#1}%
  \colorlet{ZedColor}{#1}%
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% DEFINITIONS TO MARK CHANGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\usepackage{tocloft}
\usepackage{fmtcount}
\usepackage[]{imakeidx}

\newif\ifShowChanges
\ShowChangesfalse

\makeindex[title=Index of Changes, name=changes, columns=1]
\newlistof{changes}{s}{\listchangename}

\providecolor{added}{rgb}{0,0.5,0}
\providecolor{changed}{rgb}{0.5,0.5,0}
\providecolor{deleted}{rgb}{0.5,0,0}
\providecolor{moved}{rgb}{0,0,0.5}

\newcommand{\added}[1]{{\ifShowChanges\setZColors{added}\color{added}{}\fi#1}}
\newcommand{\changed}[1]{{\ifShowChanges\setZColors{changed}\color{changed}{}\fi#1}}
\newcommand{\deleted}[1]{{\ifShowChanges\setZColors{deleted}\color{deleted}\sout{#1}\fi}}
\newcommand{\moved}[1]{{\ifShowChanges\setZColors{moved}\color{moved}{}\fi#1}}

%\newcommand{\BH}[2]{\large\textbf{\hyperlink{#1}{#2\textsuperscript{#1}}}\normalsize}
%\newcommand{\xindex}[2]{\expandafter\index\expandafter[changes]\expandafter{\pageref{#1}#2}}
\newcommand{\erasenumber}[1]{}
\newrobustcmd{\removecomma}[1]{}

\makeatletter
\newcommand{\IfFirstOccurrence}[1]{%
  \ifthmt@thisistheone%
  #1%
  \fi%
}
\makeatother
\newcommand{\addchange}[1]{%
  \ifShowChanges%
  \refstepcounter{changes}%
  {\normalsize%
    [Change \arabic{changes}]%
  }
  \phantomsection%
  \label{change\arabic{changes}}%
  \index[changes]{\expandafter\sortgref\arabic{changes}\sortgref@{%
      \hyperref[change\arabic{changes}]{%
        [Change \arabic{changes}]%
        Page \pageref{change\arabic{changes}}%
      }: {#1}}\removecomma|erasenumber}%
%  \xindex{\arabic{changes}}{@\arabic{changes}: #1}
%  \index[changes]{Comment \textbf{C\expandafter\sortgref#1\sortgref}|BH{\arabic{changes}}}%
  \fi%
}%


\def\sortgref#1\sortgref{%
	\ignoresort{\ifnum#1<10 00\else\ifnum #1<100 0\fi\fi#1}#1%
}
\protected\def\ignoresort#1{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usetikzlibrary{graphs}

\setcounter{secnumdepth}{4}

\title{\protect\addchange{Changed title to not refer to
    \protect``verification\protect''}\protect\changed{Ahead-of-Time\\Algebraic
    Compilation for\\Safety-Critical Java}} \author{James Baxter}
\date{}

\bibliography{literature} 

\newif\ifFullModel

%TC:macro \IfFullModel [ignore]
%TC:macro \IfNotFullModel [text]
\FullModelfalse

\newcommand{\IfFullModel}[1]{\ifFullModel #1 \fi}
\newcommand{\IfNotFullModel}[1]{\ifFullModel \else #1 \fi}

\newif\ifExtended

%TC:macro \IfExtended [ignore]
%TC:macro \IfNotExtended [text]
\Extendedfalse

\newcommand{\IfExtended}[1]{\ifExtended #1 \fi}
\newcommand{\IfNotExtended}[1]{\ifExtended \else #1 \fi}

\newcommand{\directsubclass}{\prec_{\mathrm{d}}}
\newcommand{\directsubclassid}[1]{\prec_{\mathrm{D},#1}}
\newcommand{\subclassid}[1]{\preceq_{#1}}

\makeatletter
\def\@endtheorem{\endtrivlist}
\def\thm@space@setup{%
  \thm@preskip=\parskip \thm@postskip=0pt
}
\makeatother

% \theoremstyle{plain}
% \newtheorem{thm}{Theorem}[section]
% \newtheorem{lem}[thm]{Lemma}
% \newtheorem{rule}{Rule}
\declaretheorem[
name=Theorem,
numberwithin=section,
refname={Theorem, Theorems},
Refname={Theorem, Theorems}
]{thm}
\declaretheorem[
name=Lemma,
numberlike=thm,
refname={Lemma, Lemmas},
Refname={Lemma, Lemmas}
]{lem}

\declaretheoremstyle[
style=plain,
notebraces={[}{]}
]{rule}

\declaretheorem[
name=Rule,
style=rule,
numbered=no,
refname={Rule,Rules},
Refname={Rule,Rules}
]{crule}

\declaretheorem[
name=Law,
style=rule,
numbered=no,
refname={Law,Laws},
Refname={Law,Laws}
]{law}

\makeatletter
\declaretheoremstyle[
style=definition,
bodyfont={
  \addtolength{\@totalleftmargin}{0.5cm}
  \addtolength{\linewidth}{-0.5cm}
  \parshape 1 0.5cm \linewidth
}
]{idefinition}
\makeatother

\declaretheorem[
name=Definition,
refname={definition,definitions},
Refname={Definition,Definitions}, numberlike=thm, style=idefinition
]{defn}

\declaretheoremstyle[
style=definition,
notebraces={[}{]}
]{proof}

\declaretheorem[
name=Proof,
style=proof,
numbered=no,
prefoothook={\qedsymbol\relax}
]{crproof}

\renewcommand{\circblockbegin}{\left(\begin{array}{l}}
\renewcommand{\circblockend}{\end{array}\right)}

\newcommand{\Apply}[1]{{\bf apply} {#1}}
\newcommand{\ApplyFor}[2]{{\bf apply} {#1}(#2)}
\newcommand{\ApplyTo}[2]{{\bf apply} {#1} {\bf to} {#2}}
\newcommand{\ApplyToFor}[3]{{\bf apply} {#1}(#3) {\bf to} {#2}}
\newcommand{\ApplyReverse}[1]{{\bf apply} {#1} {\bf in reverse}}
\newcommand{\ApplyReverseFor}[2]{{\bf apply} {#1}(#2) {\bf in reverse}}
\newcommand{\ApplyReverseTo}[2]{{\bf apply} {#1} {\bf in reverse to} {#2}}
\newcommand{\ApplyReverseToFor}[3]{{\bf apply} {#1}(#3) {\bf in reverse to} {#2}}
\newcommand{\ExhaustivelyApply}[1]{{\bf exhaustively apply} {#1}}
\newcommand{\ExhaustivelyApplyFor}[2]{{\bf exhaustively apply} {#1}(#2)}
\newcommand{\ExhaustivelyApplyTo}[2]{{\bf exhaustively apply} {#1} {\bf to} {#2}}
\newcommand{\ExhaustivelyApplyToFor}[3]{{\bf exhaustively apply} {#1}(#3) {\bf to} {#2}}
\newcommand{\ExhaustivelyApplyReverse}[1]{{\bf exhaustively apply} {#1} {\bf in reverse}}
\newcommand{\ExhaustivelyApplyReverseFor}[2]{{\bf exhaustively apply} {#1}(#2) {\bf in reverse}}
\newcommand{\ExhaustivelyApplyReverseTo}[2]{{\bf exhaustively apply} {#1} {\bf in reverse to} {#2}}
\newcommand{\ExhaustivelyApplyReverseToFor}[3]{{\bf exhaustively apply} {#1}(#3) {\bf in reverse to} {#2}}
\algloopdefx{Match}[2]{{\bf match} #1 {\bf with} #2 {\bf then}}
\algloopdefx{MatchIn}[2]{{\bf match} #1 {\bf in} #2 {\bf then}}
\algloopdefx{MatchThen}[1]{{\bf match} #1 {\bf then}}
\algloopdefx{Let}[1]{{\bf let} #1 {\bf in}}
\newcommand{\Matches}[2]{#1 {\bf matches} #2}
\algblockx[Try]{Try}{EndTry}
  {{\bf try}}
  {{\bf end try}}
\algblockx[ExhaustivelyTry]{ExhaustivelyTry}{EndExhaustivelyTry}
  {{\bf exhaustively try}}
  {{\bf end try}}
\algblockx[MatchInBlock]{MatchInBegin}{EndMatch}
  [2]{{\bf match} #1 {\bf in} #2 {\bf then begin}}
  {{\bf end match}}

\MakeRobust{\Call}
\LetLtxMacro{\OldCall}{\Call}
\renewcommand{\Call}[2]{\OldCall{\mbox{#1}}{#2}}
\MakeRobust{\Let}

%TC:group zed 0 displaymath
%TC:group axdef 0 displaymath
%TC:group schema 1 displaymath
%TC:group circus 0 displaymath
%TC:group circusaction 0 displaymath
%TC:macroword \Circus 1
%TC:group theorem 1 displaymath
%TC:group zproof 0 displaymath 
%TC:group algorithm 0 float
%TC:group argue 0 displaymath
  
\begin{document}

\begin{titlepage}
  \centering{}
  \hfill
  \vfill
  \vfill
  \vfill
  {\Huge\bf \thetitle}
  \vfill
  \vfill
  {\huge \theauthor}
  \vfill
  \vfill
  {\LARGE PhD}
  \vfill
  {\LARGE University of York}
  \vfill
  {\LARGE Computer Science}
  \vfill
  {\LARGE July 2018}
\end{titlepage}

\null
\thispagestyle{empty}%
\addtocounter{page}{-1}%
\newpage

\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{Abstract}
\begin{abstract}
  \thispagestyle{plain}
  In recent years Java has been increasingly considered as a language
  for safety-critical embedded systems.
  However, some features of Java are unsuitable for such systems.
  This has resulted in the creation of Safety-Critical Java (SCJ),
  which facilitates the development of certifiable real-time and
  embedded Java programs.
  SCJ uses different scheduling and memory management models to
  standard Java, so it requires a specialised virtual machine (SCJVM).
  A common approach is to compile Java bytecode program to a native
  language, usually C, ahead-of-time for greater performance on
  low-resource embedded systems.
  
  Given the safety-critical nature of the applications, it must be
  ensured that the virtual machine is correct.
  However, so far, formal verification has not been applied to any
  SCJVM.
  This thesis contributes to the formal verification of SCJVMs that
  utilise ahead-of-time compilation by presenting a verification of
  compilation from Java bytecode to C.
  
  The approach we adopt is an adaptation of the algebraic approach
  developed by Sampaio and Hoare.
  We start with a formal specification of an SCJVM executing the
  bytecodes of a program, and transform it, through the application
  of proven compilation rules, to a representation of the target C
  code.
  Thus, our contributions are a formal specification of an SCJVM, a
  set of compilation rules with proofs, and a strategy for applying
  those compilation rules.
  
  Our compilation strategy can be used as the basis for an
  implementation of an ahead-of-time compiling SCJVM, or verification
  of an existing implementation.
  Additionally, our formal model of an SCJVM may be used as a
  specification for creating an interpreting SCJVM.
  To ensure the applicability of our results, we base our work on
  icecap, the only currently available SCJVM that is open source and
  up-to-date with the SCJ standard.
\end{abstract}

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Contents}
\tableofcontents

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures

% acknowledgements page
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgements}
\chapter*{Acknowledgements}

Firstly, I would like to thank my supervisor, Ana Cavalcanti, for all
her advice on the work and comments on this thesis as it developed,
and for always replying to emails promptly when I required help.
My thanks also go to Andy Wellings, for the clarifications of the
Safety Critical Java specification that he offered and for bringing
the issues I raised to the attention of the expert group developing
that specification.
I would also like to thank my assessor, Colin Runciman, for his
comments on drafts of the thesis and suggestions of related
literature.

Leo Freitas gave great help with the use of Z/EVES and Community Z
Tools, and I am thankful to him for the time he took to explain these
things.
My thanks also go to the authors of the icecap HVM for their help in
using it.
I am thankful to Augusto Sampaio for his feedback on my use of the
algebraic approach.
I would also like to thank Pedro Ribeiro for his useful advice on the
format of this thesis, and Alvaro Miyazawa for his help in putting
technical reports online.
Heartfelt thanks go to Simon Foster for all his friendship throughout
this PhD, for teaching me to use the Isabelle proof assistant, and for
introducing me to the area of formal methods, without which I may
never have done this PhD at all.

This work has been funded by an EPSRC studentship, and I am grateful to
EPSRC for their provision of the funding and to the University for
deciding to allocate one of the studentships to me.

My parents have given me a great amount of support and encouragement
over the years, and I am particularly grateful to them for how they
have supported me while I have been busy writing this thesis. 
I would particularly like to thank my mother for proofreading this
thesis.
My thanks also go to all my friends for their support, particularly to
the members of Trinity Church York and Wentworth Christian Union for
all their support, encouragement, and prayers.

Finally, as a Christian, I acknowledge that all that happens is in the
hands of God. 
I am enormously thankful to him for all he has done in bringing me
safely to this point. 
Therefore, I commit this thesis to his sovereign care and ask that he
alone may be glorified in all things.

% end acknowledgements page

% declaration page
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Declaration}
\chapter*{Declaration}

I declare that this thesis is a presentation of original work and I am
the sole author. 
This work has not previously been presented for an award at this, or
any other, University. 
All sources are acknowledged as references.
The following material, presented in this thesis, has been previously
published:

\begin{refsection}
  \raggedright
  \nocite{baxter2015a,baxter2017}
  \setlength{\bibitemsep}{0.5cm}
  \printbibliography[resetnumbers=true,heading=none]
\end{refsection}

The following published material is not directly described in this
thesis, but provides evidence of the context in which our work can be
applied. We discuss it here.

\begin{refsection}
  \raggedright
  \nocite{freitas2016}
  \setlength{\bibitemsep}{0.5cm}
  \printbibliography[resetnumbers=3,heading=none]
\end{refsection}

% end declaration page
\begin{refsection}

\input{introduction.tex}

\input{field-survey.tex}

\input{scjvm-services.tex}

\input{cee.tex}

\input{compilationstrategy.tex}

\input{evaluation.tex}

\input{conclusion.tex}

\appendix

%TC:macro \IfFullModel [text]
%TC:macro \IfNotFullModel [ignore]
\FullModeltrue
\chapter{Compilation Rules}
\label{compilation-rules-appendix}

\input{compilationrules.tex}

\chapter{C Code of Examples}
\label{example-c-code-appendix}

\input{examplecode.tex}

{\raggedright \printbibliography[resetnumbers=true]}
\end{refsection}

\ifShowChanges
\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{Index of Changes}
\printindex[changes]
\fi

\end{document}
